Supp mat: time series paper

### 2017-06-05 writing/manuscript_draft.Rmd
changes to figures and supp mat

#Load our cleaned data (see other code for cleaning steps)
  load('cleaned-data/pop_info_example_goodNAs.Rdata')
```

### Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]`) and were left with `r nrow(pop_info)` time. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

@@

We evaluated the question of what minimum time series is required to answer a question of interest. More specifically we ask, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance? The minimum time series length required is that which had high enough statistical power. We investigated this question through two approaches: simulation-based and empirical analyses. In both cases, we found the minimum time series length required depended strongly on the trend strength, population variability, and autocorrelation.

@@

@@

We show two different approaches to estimate the minimum time series length required required to address a particular question of interest. The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987]. Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.

@@

### Past work on this issue
There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required. We show that to assess longterm trends in abundance, 10-15 years of continuos monitoring are required. In line with theoretical predictions, we also show that $t_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation. Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of longterm monitoring programs. (where do IUCN recommendations come from?) From both a scientific and management perspective estimates of $t_{min}$ are important. If a time series is too short, we are likely to make incorrect conclusions about longterm trends. In addition, a time series that is too long may have been a poor use of scare funds. Future work can examine other species, with a wider range of life history characteristics, or address other questions besides longterm trends in abundance.

### Time series length recommendations from other papers

- greater than 5 years [@Morrison2008]
- more than 10-15 years [@Rueda-Cediel2015]
- general advice (variability, correlations) [@Rhodes2011]
- more than 10 years [@Gerber1999]
- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
- paper on syncrony from James

>>>>

#Load our cleaned data (see other code for cleaning steps)
  load('cleaned-data/pop_info_example_goodNAs.Rdata') #Do i need to filter out non-significant populations as well (where no trend was present)
```

### Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data is originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]`), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, longterm trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterian A2 to classify a species as vulnerable [@IUCN2012].

@@

We evaluated the question of what minimum time series is required to answer a question of interest. More specifically we asked, what is the minimum time series length required to determine, via linear regression, the longterm trend in population abundance? The minimum time series length required is that which had high enough statistical power. We investigated this question through two approaches: simulation-based and empirical analyses. In both cases, we found the minimum time series length required depended strongly on the trend strength, population variability, and autocorrelation.

@@

### Questions in the appendix

Describe some of the results that we include in the appendix

@@

We show two different approaches to estimate the minimum time series length required required to address a particular question of interest. The simulation-based approach has been suggested by others, especially in situations where analysis is more complicated than a simple power analysis [@Gerrodette1987]. These simulations support past work...

Here, we focus on a empirical approach to estimate the minimum time series length requried to address a given question.

@@

### Past work on this issue - can I integrate this better into the rest of the section. How much should go in introduction?
There have been previous attempts to address how long a time series should be to assess trends in abundance. For example,...

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required. We show that to assess longterm trends in abundance, 10-15 years of continuos monitoring are required (Fig. \ref{fig:min_time_dist}. In line with theoretical predictions, we also show that $t_{min}$ is strongly correlated with the overall trend in abundance, variability in abundance, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of longterm monitoring programs. (where do IUCN recommendations come from?) From both a scientific and management perspective estimates of $t_{min}$ are important. If a time series is too short, we are likely to make incorrect conclusions about longterm trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work can examine other species, with a wider range of life history characteristics, or address other questions besides longterm trends in abundance.

<!--
### Time series length recommendations from other papers

- greater than 5 years [@Morrison2008]
- more than 10-15 years [@Rueda-Cediel2015]
- general advice (variability, correlations) [@Rhodes2011]
- more than 10 years [@Gerber1999]
- specific recommendations depend on amount of power and strength of trend (Bart 2004, Hatch 2003) [@Seavy2007]
- paper on syncrony from James
-->

### 2017-06-05 writing/manuscript_draft.tex
changes to figures and supp mat

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35) and were left with 1019 time. The
database includes information on 499 vertebrate species with a focus on
mammals, birds, and fish. The database also includes information on
generation length and census specifications. For each time series, we
also calculated variables of interest like variance in population size,
longterm trend in abundance (slope coefficient from simple linear
regression), and temporal autocorrelation. Lastly, we also calculated
the minimum number of years required (10 years or 3 generations) under
IUCN Red List Criterian A2 to classify a species as vulnerable (IUCN
2012).

@@

@@

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). Here, we focus on a empirical approach to
estimate the minimum time series length requried to address a given
question.

@@

\subsubsection{Past work on this issue}\label{past-work-on-this-issue}

@@

We use a database of 868 populations to determine the minimum time
series length required. We show that to assess longterm trends in
abundance, 10-15 years of continuos monitoring are required. In line
with theoretical predictions, we also show that \(t_{min}\) is strongly
correlated with the overall trend in abundance, variability in
abundance, and the temporal autocorrelation. Our work implies that for
many populations, census data less than 10-15 years are probably not
reliable. This stresses the importance of longterm monitoring programs.
(where do IUCN recommendations come from?) From both a scientific and
management perspective estimates of \(t_{min}\) are important. If a time
series is too short, we are likely to make incorrect conclusions about
longterm trends. In addition, a time series that is too long may have
been a poor use of scare funds. Future work can examine other species,
with a wider range of life history characteristics, or address other
questions besides longterm trends in abundance.

\subsubsection{Time series length recommendations from other
papers}\label{time-series-length-recommendations-from-other-papers}

\begin{itemize}
\tightlist
\item
  greater than 5 years (Morrison and Hik 2008)
\item
  more than 10-15 years (Rueda-Cediel et al. 2015)
\item
  general advice (variability, correlations) (Rhodes and Jonzen 2011)
\item
  more than 10 years (Gerber, DeMaster, and Kareiva 1999)
\item
  specific recommendations depend on amount of power and strength of
  trend (Bart 2004, Hatch 2003) (Seavy and Reynolds 2007)
\item
  paper on syncrony from James
\end{itemize}

>>>>

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data is originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35), and those with missing data, and were
left with 1019 time series. The database includes information on 499
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, longterm trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation. Lastly, we also calculated the minimum number of years
required (10 years or 3 generations) under IUCN Red List Criterian A2 to
classify a species as vulnerable (IUCN 2012).

@@

\subsubsection{Questions in the
appendix}\label{questions-in-the-appendix}

Describe some of the results that we include in the appendix

@@

We show two different approaches to estimate the minimum time series
length required required to address a particular question of interest.
The simulation-based approach has been suggested by others, especially
in situations where analysis is more complicated than a simple power
analysis (Gerrodette 1987). These simulations support past work\ldots{}

Here, we focus on a empirical approach to estimate the minimum time
series length requried to address a given question.

@@

\subsubsection{Past work on this issue - can I integrate this better
into the rest of the section. How much should go in
introduction?}\label{past-work-on-this-issue---can-i-integrate-this-better-into-the-rest-of-the-section.-how-much-should-go-in-introduction}

@@

We use a database of 868 populations to determine the minimum time
series length required. We show that to assess longterm trends in
abundance, 10-15 years of continuos monitoring are required (Fig.
\ref{fig:min_time_dist}. In line with theoretical predictions, we also
show that \(t_{min}\) is strongly correlated with the overall trend in
abundance, variability in abundance, and the temporal autocorrelation
(Fig. \ref{fig:correlates}). Our work implies that for many populations,
census data less than 10-15 years are probably not reliable. This
stresses the importance of longterm monitoring programs. (where do IUCN
recommendations come from?) From both a scientific and management
perspective estimates of \(t_{min}\) are important. If a time series is
too short, we are likely to make incorrect conclusions about longterm
trends. In addition, a time series that is too long may be a poor use of
already limited funds (Gerber, DeMaster, and Kareiva 1999). Future work
can examine other species, with a wider range of life history
characteristics, or address other questions besides longterm trends in
abundance.

### 2017-06-05 writing/supp_mat_draft.Rmd
changes to figures and supp mat

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
    citation_package: natbib
fontsize: 12pt
geometry: margin=0.75in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \renewcommand{\thepage}{S\arabic{page}} 
    \renewcommand{\thesection}{S\arabic{section}}  
    \renewcommand{\thetable}{S\arabic{table}}  
    \renewcommand{\thefigure}{S\arabic{figure}}
---

@@

Dec XX, 2016
 \end{center}

@@

1. Mathematical description of our model
2. Inverse pattern-orientated estimation procedure

Code for all the figures and tables can be found on [Github](https://github.com/erwhite1).

@@

```{r, echo=F, fig.cap='duh\\label{fig:growth_rate}'}
#Example of using population growth rate instead of power
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
source("scripts/calculate_geometric_growth.R")
load('cleaned-data/pop_info_example_goodNAs.Rdata')

par(mfrow=c(1,2))
  pop = subset(long_dat,long_dat$ID=='10696') #13192 for Harp seals
  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,ylim=c(0,2000),las=1)
    abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = 'year',1,line=3,cex=1.4)
    mtext(text = 'population size',2,line=3.5,cex=1.4)

  source('scripts/create_subsamples.R')
  aaa=calculate_average_geometric_growth(pop$popvalue)
  plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
    mtext(text = 'years sampled',1,line=3,cex=1.4)
    mtext(text = 'population growth rate',2,line=3.7,cex=1.4)
```

@@

TEXT HERE ON IDEA

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
# show gam example side-by-side plot
```

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
# show gam results for minimium time required
```

### References

>>>>

---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: true
    citation_package: natbib
fontsize: 12pt
geometry: margin=1in
bibliography: White_bib.bib
header-includes: 
    \usepackage{float}
    \renewcommand{\thepage}{S\arabic{page}} 
    \renewcommand{\thesection}{S\arabic{section}}  
    \renewcommand{\thetable}{S\arabic{table}}  
    \renewcommand{\thefigure}{S\arabic{figure}}
---

@@

June 22, 2017
 \end{center}

@@

1. Detailed example of subsampling and power calculations
2. Figures and results from other questions of interest
3. Additional figures?
4. Additional references

Code for all the figures and tables can be found at [](https://github.com/erwhite1).

@@

Instead of detecting a trend over time with linear regression, we could also calculate the geometric growth rate of the population. In figure \ref{fig:growth_rate}, we show how to calculate growth rates for subsamples of a time series. First, we created subsamples of each possible length from the full 35 year time series, as we did in the main next. Next, we calculated the mean and standard deviation of growth rates for each possible time series length (Fig. \ref{fig:growth_rate}b). Lastly, we calculated the percent error $\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|$ between the mean of each time series length (observed) and the overall population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c), we show the percent error as a function of time series length. Here we define the minimum time required as the minimum number of years to achieve less than 20% error.

```{r, echo=F, fig.cap='Example of calculating minimum time required for growth rate estimation. (a) abundance over time, (b) mean and standard deviation of growth rate for subsamples of entire time series, and (c) the percent error between mean estimated growth rate and the true longterm growth rate. The vertical bar denotes the minimum time required to estimate growth rate within 10% error.\\label{fig:growth_rate}'}
#Example of using population growth rate instead of power
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
source("scripts/calculate_geometric_growth.R")
load('cleaned-data/pop_info_example_goodNAs.Rdata')

par(mfrow=c(1,3),oma=c(0,1,0,0))

  pop = subset(long_dat,long_dat$ID=='10007') #13192 for Harp seals
  source('scripts/create_subsamples.R')
  source('scripts/percent_error.R')
  aaa=calculate_average_geometric_growth(pop$popvalue)
  plot(pop$year,pop$popvalue,cex.axis=1.2,cex.lab=1.2,ylab='',xlab='',pch=16,las=1)
    #abline(lm(pop$popvalue~pop$year),lwd=2,col='red')
    mtext(text = 'year',1,line=3,cex=1.4)
    mtext(text = 'population size',2,line=3.5,cex=1.4)

  plot(rowMeans(aaa,na.rm=T),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
    mtext(text = 'years sampled',1,line=3,cex=1.4)
    mtext(text = 'population growth rate',2,line=3.7,cex=1.4)

    abline(h= -0.01*rowMeans(aaa,na.rm = T)[34] +rowMeans(aaa,na.rm = T)[34] )
    abline(h= 0.01*rowMeans(aaa,na.rm = T)[34] +rowMeans(aaa,na.rm = T)[34] )

    arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)+sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)
    arrows(1:34,rowMeans(aaa,na.rm=T),1:34,rowMeans(aaa,na.rm=T)-sd(apply(aaa,MARGIN = 1,FUN=var,na.rm=T),na.rm=T),angle = 90,length = 0.1)

    plot(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34]),cex.axis=1.2,pch=16,ylab='',xlab='',las=1)
    mtext(text = 'years sampled',1,line=3,cex=1.4)
    mtext(text = 'percent error in growth rate',2,line=3.7,cex=1.4)
    abline(h=0.1,lty=2)
    abline(v=tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.1),1)+1,lty=2)
```

We applied the same calculations to `r nrow(pop_info)` populations, as we did in the main text. We then obtain a distribution of the minimum time required to measure the ``true" longterm growth rate (Fig. \reg{fig:min_time_growth_dist}). We see a bimodal distribution with many populations required 30+ years to estimate the longterm growth rate. The large number of short year required is due to cases where the entire time series is consistently increasing or decreasing at the same rate.

```{r, echo=F,fig.cap='Histogram of the minimum time required in order to estimate the longterm growth rate within 20% error.\\label{fig:min_time_growth_dist}'}
# Growth rates calculations for each species
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load('cleaned-data/pop_info_example_goodNAs.Rdata')

 for (j in 1:nrow(pop_info)){
    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:35,]
    pop$popvalue=as.numeric(pop$popvalue)
    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way

   # source('scripts/calculate_power_metric.R')
    aaa=calculate_average_geometric_growth(pop$popvalue)
    if (length(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2))==0){
      pop_info$min_time_growth_rate[j] = 2
    }else{
      pop_info$min_time_growth_rate[j] = tail(which(percent_error(rowMeans(aaa,na.rm=T),rowMeans(aaa,na.rm=T)[34])>0.2),1)+1}
    #print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
 }

# Create plot
par(mfrow=c(1,1))
hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
mtext(text = 'Frequency',2,line=3,cex=1.2)
mtext(text = 'Minimum time required',1,line=3,cex=1.2)

```

\clearpage

@@

In the main text, we examined the minimum time required to identify a trend in abundance via linear regression. This approach allowed us to identify increases or decreases, but a linear model may not always be a good fit. Generalized additive models (GAMs) are more general than general linear models and allow more flexible. GAMs are models where a response variable depends on unknown smooth functions of explanatory variables. GAMs, therefore, can identify relationships between response variables and explantory variables that are non-linear and perhaps more complicated. The downsite of GAMs...

Here, we conduct the same analyses in the main text, but instead calculate the minimum time series required to detect trends over time according to a GAM model. We hypothesize that GAMs should require less time to detect a trend as they are more flexible than linear regression. We provide an example that shows statistical power increases with more time sampled (Fig. \ref{fig:gam_example}).

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_example}'}
# show gam example side-by-side plot
plot(1,1)
```

We then fit GAM models for XX populations. We found... (Fig. \ref{fig:gam_result}). This is shorter/longer than the main text question.

```{r,echo=F,fig.cap='results here from GAM.\\label{fig:gam_result}'}
# show gam results for minimium time required
setwd("/Users/eastonwhite/Desktop/Research/time-series-project")
load('cleaned-data/pop_info_example_goodNAs.Rdata')

 for (j in 1:nrow(pop_info)){
    pop= subset(long_dat,long_dat$ID==pop_info$ID[j])[1:35,]
    pop$popvalue=as.numeric(pop$popvalue)
    pop$popvalue=pop$popvalue/max(pop$popvalue) #should I standaridize this in a better way

    source('scripts/calculate_power_metric_gam.R')
    pop_info$min_time_for_power_gam[j] = min_time_needed(pop$popvalue,0.05,0.8)
    print(paste(j,':',pop_info$min_time_for_power[j],sep=' '))
 }

# Create plot
par(mfrow=c(1,1))
hist(pop_info$min_time_growth_rate,las=1,ylab='',xlab='',cex.axis=1.2,main='',breaks=20,xlim=c(0,40),ylim=c(0,500))
mtext(text = 'Frequency',2,line=3,cex=1.2)
mtext(text = 'Minimum time required',1,line=3,cex=1.2)

```

### Minimum time to detect density-dependence in time series

For decades, the importance of density-dependence in natural populations has been an important question in ecology (citations). To address this question, @Knape2012 used data from the Global Population Dynamics Database [@GPDD2010]. They fitted 627 time series with a Gompertz population model and found evidence for density-dependence in 45% of the time series. Each time series in their database contained at least 15 datapoints, and ranged from 15 to 157. A follow up question to their work would be, what time series length is required to confidently estimate density-dependence.

\clearpage
### References

### 2017-06-05 writing/supp_mat_draft.tex
changes to figures and supp mat

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=0.75in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mathematical description of our model
\item
  Inverse pattern-orientated estimation procedure
\end{enumerate}

Code for all the figures and tables can be found on
\href{https://github.com/erwhite1}{Github}.

@@

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{duh\label{fig:growth_rate}}
\end{figure}

@@

TEXT HERE ON IDEA

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{plainnat}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Detailed example of subsampling and power calculations
\item
  Figures and results from other questions of interest
\item
  Additional figures?
\item
  Additional references
\end{enumerate}

Code for all the figures and tables can be found at
\href{https://github.com/erwhite1}{}.

@@

Instead of detecting a trend over time with linear regression, we could
also calculate the geometric growth rate of the population. In figure
\ref{fig:growth_rate}, we show how to calculate growth rates for
subsamples of a time series. First, we created subsamples of each
possible length from the full 35 year time series, as we did in the main
next. Next, we calculated the mean and standard deviation of growth
rates for each possible time series length (Fig.
\ref{fig:growth_rate}b). Lastly, we calculated the percent error
\(\mbox{percent error} = 100 \times \left| \frac{\mbox{observed} - \mbox{theoretical}}{\mbox{theoretical}} \right|\)
between the mean of each time series length (observed) and the overall
population growth rate (theoretical). In Fig. \ref{fig:growth_rate}c),
we show the percent error as a function of time series length. Here we
define the minimum time required as the minimum number of years to
achieve less than 20\% error.

@@

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{Example of calculating minimum time required for growth rate
estimation. (a) abundance over time, (b) mean and standard deviation of
growth rate for subsamples of entire time series, and (c) the percent
error between mean estimated growth rate and the true longterm growth
rate. The vertical bar denotes the minimum time required to estimate
growth rate within 10\% error.\label{fig:growth_rate}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{Histogram of the minimum time required in order to estimate the
longterm growth rate within 20\% error.\label{fig:min_time_dist}}
\end{figure}

@@

In the main text, we examined the minimum time required to identify a
trend in abundance via linear regression. This approach allowed us to
identify increases or decreases, but a linear model may not always be a
good fit. Generalized additive models (GAMs) are more general than
general linear models and allow more flexible. GAMs are models where a
response variable depends on unknown smooth functions of explanatory
variables. GAMs, therefore, can identify relationships between response
variables and explantory variables that are non-linear and perhaps more
complicated. The downsite of GAMs\ldots{}

Here, we conduct the same analyses in the main text, but instead
calculate the minimum time series required to detect trends over time
according to a GAM model. We hypothesize that GAMs should require less
time to detect a trend as they are more flexible than linear regression.
We provide an example that shows statistical power increases with more
time sampled (Fig. \ref{fig:gam_example}).

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{results here from GAM.\label{fig:gam_example}}
\end{figure}

We then fit GAM models for XX populations. We found\ldots{} (Fig.
\ref{fig:gam_result}). This is shorter/longer than the main text
question.

\begin{figure}[htbp]
\centering
\includegraphics{supp_mat_draft_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{results here from GAM.\label{fig:gam_result}}
\end{figure}

\subsubsection{Minimum time to detect density-dependence in time
series}\label{minimum-time-to-detect-density-dependence-in-time-series}

For decades, the importance of density-dependence in natural populations
has been an important question in ecology (citations). To address this
question, \citet{Knape2012} used data from the Global Population
Dynamics Database \citep{GPDD2010}. They fitted 627 time series with a
Gompertz population model and found evidence for density-dependence in
45\% of the time series. Each time series in their database contained at
least 15 datapoints, and ranged from 15 to 157. A follow up question to
their work would be, what time series length is required to confidently
estimate density-dependence.
